Automated iterative forecasting for the Portal Project

### 2018-02-02 manuscript/portal_iterative_forecasting_ms.Rmd
Vertical figure 1 a and b, reorder figure references in ms

Implementation of iterative forecasting requires the regular rebuilding of models with new raw data as it becomes available and the presentation of those forecasts in usable forms; in our case, this occurs monthly. Rebuilding models in an efficient and maintainable way relies on developing an automated pipeline to handle the six stages of converting raw data into new forecasts: data collection, data sharing, data manipulation, modeling and forecasting, archiving, and presenting of the forecasts (Figure 1). To implement the pipeline outlined in Figure 1, we used a "continuous analysis" framework [*sensu* @beaulieu2017] that automatically processes the most up-to-date data, refits the models, makes new forecasts, archives the forecasts, and updates a website with analysis of current and previous forecasts. In this section we describe our approach to streamlining and automating the multiple components of the forecasting pipeline and the tools and infrastructure we employed to execute each stage of the pipeline.

![Figure 1. Stages of the forecasting pipeline. To go from raw data to forecast presentation involves a number of stages, each of which requires unique tasks, tools and infrastructure. While each stage has unique tasks associated with it, the stages of our pipeline are also interdependent. Our pipeline is primarily a linear structure where outputs from one stage form the inputs for the subsequent stage. All stages require code written in R. ](figs/pipeline_fig.png).

@@

A core aspect of iterative near-term forecasting is the regular rerunning of the forecasting pipeline. We chose to have the computer run our pipeline and employed "continuous analysis" [*sensu* @beaulieu2017] to drive the automation of both the full pipeline and a number of its individual components. Continuous analysis uses a set of tools originally designed for software development called "continuous integration" (CI). CI combines computing environments for running code with monitoring systems to identify changes in data or code. Essentially, CI is a computer helper whose job is to watch the pipeline and, when it sees a change in the code or data, it runs all the computer scripts needed to ensure that the forecasting pipeline runs from beginning to end. This is useful for iterative near-term forecasting because it does not rely on humans to remember to create forecasts when new models or data are added. These tools are common in the area of software development where they are used to automate software testing and integrate work by multiple developers working on the same code base. However, these tools can be used for any computational task that needs to be regularly repeated or run after changes to code or data [@beaulieu2017]. Our forecasting pipeline currently runs on a publicly available continuous integration service (Travis CI; https://travis-ci.org/) that is free for open source projects (up to a limited amount of computing time). Because of the widespread use of CI in software development, alternative services that can run code on local or cloud-based computational infrastructure also exist (e.g., Drone (http://try.drone.io/)) [@beaulieu2017]. As detailed below, we use CI to quality check data, test code using "unit tests" [@wilson2014], build models, make forecasts, and publicly present and archive the results (Figure 2).

![Figure 2. Continuous integration system. Each box denotes the core infrastructure used for each stage of the forecasting pipeline denoted in Figure 1. Continuous integration, denoted here with the Travis icon (a girl wearing safety glasses and hardhat), plugs into our pipeline at every stage.  Travis triggers the code involved in major events that involve integration across stages of the pipeline, such as taking the output from the forecasting stage (purple box) to create an updated presentation (rose box). Travis also runs tasks within a stage, such as conducting tests to make sure code changes have not introduced errors (icons on arrows originating and ending on the same box)](figs/contanaly_fig.png)

@@

Iterative forecasting benefits from frequently updated data so that state changes can be quickly incorporated into new forecasts [@dietze2018]. Frequent data collection and rapid processing are both important for providing timely forecasts. Since our data is collected monthly, ensuring that the models have access to the newest data requires a data latency period of less than 1 month from collection to availability for modeling. To accomplish this, we automated components of the data processing and quality assurance/quality control (QA/QC) process to reduce the time needed to add new data to the database (Figures 1 and 2).

@@

Once data is available, it needs to be processed into a form appropriate for modeling (Figure 1). For many ecological datasets, this requires not only simple data manipulation but also a good understanding of the data to facilitate appropriate aggregation. Data manipulation steps are often conducted using custom one-off code to convert the raw data into the desired form [@morris2013], but this approach has several limitations. First, each researcher must develop and maintain their own data manipulation code, which is inefficient and can result in different researchers producing different versions of the data for the same task. Subtle differences in data processing decisions have led to confusion when reproducing results for the Portal data in the past. Second, this kind of code is rarely robust to changes in data structure and location. Based on our experience developing and maintaining the Data Retriever [@morris2013; @senyondo2017], these kinds of changes are common. Finally, this kind of code is generally poorly tested, which can lead to errors based on mistakes in data manipulation. To avoid these issues for the Portal Project data, the Portal team has been developing an R package (portalR; http://github.com/weecology/portalr) for acquiring the data and handling common data cleaning and aggregation tasks. As a result, our modeling and forecasting code only needs to install this package and run the data manipulation and summary functions to get the appropriate data (Figure 2). The package undergoes thorough automated unit tests to ensure that data manipulations are achieving the desired results. Having data manipulation code maintained in a separate package that focuses on consistently providing properly summarized forms of the most recent data has made maintaining the forecasting code itself much more straightforward.

@@

Iterative near-term forecasting involves regularly refitting a variety of different models (Figure 1). Ideally, new models should be easy to incorporate to allow for iterative improvements to the general modeling structure and approach. We use CI to refit the models and make new forecasts each time the modeling code changes and when new data becomes available (Figure 2). We use a plugin infrastructure to allow new models to be easily added to the system. This approach treats each model as an interchangable black boxes - all models have access to the same input data and generate the same structure for model outputs (Figure 3). Each iteration of the forecasting cycle, all models that are present are run and the standardized outputs are combined into a single file to store the results of the different models' forecasts. A weighted ensemble model is then added with weights based on how well individual models fit the training data. This plugin infrastructure makes it easy to add and compare very different types of models, from the basic time-series approaches currently implemented to complex state-space and machine learning models. As long as a model script can load the provided data and produce the appropriate output it will be run and its results incorporated into the rest of the forecasting system.

![Figure 3. Demonstration of plugin infrastructure. All model scripts (represented here by the example AR1.R) are housed in a single folder. Each model script uses data provided by the core forecasting code (represented here by rodent.csv) and returns its forecast outputs in a predefined structure that is consistent across models (represented here by the example 2017_12_08forecasts.csv). Outputs from all models run on a particular date are combined into the same file (i.e. 2017_12_08forecasts.csv) to allow cross-model evaluations. Model outpus files are housed in a folder containing all forecast outputs from all previous dates to facilitate our arhciving process and and forecast assessment activities ](figs/plugin_fig.gif)

In addition to flexibility in model internals, we also wanted to support flexibility in what the models predict. Allowing models to make forecasts for system properties ranging from individual species' population abundances to total community biomass facilitates exploration of differences in forecastability across different aspects of ecological systems. We designed a forecast output format to support this. Each forecast output file contains the date being forecast, the collection date of the data used for fitting the models, the date the forecast was made, the state variable being forecast (e.g., rodent biomass, the abundance of a species), and the forecast value and associated uncertainty of that forecast (Figure 3). This allows us to store a variety of different forecasts in a common format and may serve as a useful starting point for developing a standard for storing ecological forecasts more generally.

@@

In addition to archiving the results of each forecast, we present them on a website that displays monthly rodent forecasts, model evaluation metrics, monthly reports, and information about the study site (Figure 4; [http://portal.naturecast.org](http://portal.naturecast.org/)). The website includes a graphical presentation of the most recent month's forecasts (including uncertainty) and compares the latest data to the previous forecasts. Information on the the species and the field site targeted to a general audience are also included. The site is built using Rmarkdown [@rmarkdown], which naturally integrates into the pipeline, and is automatically updated after each forecast. The `knitr` R package [@xie2015] compiles the code into HTML, which is then published using Github Pages (https://pages.github.com/). The files for the website are stored in a subdirectory of the forecasting repository. As a result, the website is also archived automatically as part of the forecast archiving.

![Figure 4. Screen capture of the homepage of the Portal Forecasting website; http://portal.naturecast.org. This site contains information on the most current forecasts, evaluation of forecast performance, and general information about the species being forecast.](figs/website_fig.png)

@@

Following the recommendations of Dietze et al [-dietze2018], we developed an automated iterative forecasting system (Figures 1 and 2) to support repeated forecasting of an ecological system. Our forecasting system automatically acquires and processes the newest data, refits the models, makes new forecasts, publicly archives those forecasts, and presents both the current forecast and information on how previous forecasts performed. Every week our forecasting system generates a new set of forecasts with no human intervention, except for the entry of new field data. This ensures that forecasts based on the most recent data are always available and allows us to rapidly assess the performance of multiple forecasting models for a number of different states of the system, including the abundances of individual species and community-level variables such as total abundance. To create this iterative near-term forecasting system, we used R to process data and conduct analyses, and we leveraged already existing services (i.e. GitHub, Travis, Docker) for more complicated cyberinfrastructure tasks. Thus, our approach to developing iterative near-term forecasting infrastructure provides an example for how short-term ecological forecasting systems can be initially developed.

We designed this forecasting system with the goal of making it relatively easy to build, maintain, and extend. We used existing technology for both running the pipeline and building individual components, which allowed us to build the system relatively cheaply in terms of both time and money. This included the use of tools like Docker for reproducibility, the Travis CI continuous integration system for automatically running the pipeline, Rmarkdown and `knitr` for generating the website, and the already existing integration between Github and Zenodo  to archive the forecasts. By using this "continuous analysis" approach [@beaulieu2017], where analyses are automatically rerun when changes are made to data, models, or associated code, we have reduced the time required by scientists to run and maintain the forecasting pipeline. To make the system extensible so that new models could be easily incorporated, we used a plugin-based infrastructure so that adding a new model to the system is as easy as adding a single file to the 'models' folder in our repository (Figure 3). This should substantially lower the barriers to other scientists contributing models to this forecasting effort. We also automatically archive the resulting forecasts publicly so that the performance of these forecasts can be assessed by both us and other researchers as new data is collected. This serves as a form of pre-registration by providing a quantitative record of the forecast before the data being predicted were collected.

>>>>

Implementation of iterative forecasting requires the regular rebuilding of models with new raw data as it becomes available and the presentation of those forecasts in usable forms; in our case, this occurs monthly. Rebuilding models in an efficient and maintainable way relies on developing an automated pipeline to handle the six stages of converting raw data into new forecasts: data collection, data sharing, data manipulation, modeling and forecasting, archiving, and presenting of the forecasts (Figure 1a). To implement the pipeline outlined in Figure 1a, we used a "continuous analysis" framework [*sensu* @beaulieu2017] that automatically processes the most up-to-date data, refits the models, makes new forecasts, archives the forecasts, and updates a website with analysis of current and previous forecasts. In this section we describe our approach to streamlining and automating the multiple components of the forecasting pipeline and the tools and infrastructure we employed to execute each stage of the pipeline.

![Figure 1. a) Stages of the forecasting pipeline. To go from raw data to forecast presentation involves a number of stages, each of which requires unique tasks, tools and infrastructure. While each stage has unique tasks associated with it, the stages of our pipeline are also interdependent. Our pipeline is primarily a linear structure where outputs from one stage form the inputs for the subsequent stage. All stages require code written in R. b) Continuous integration system. Each box denotes the core infrastructure used for each stage of the forecasting pipeline denoted in Figure 1a. Continuous integration, denoted here with the Travis icon (a girl wearing safety glasses and hardhat), plugs into our pipeline at every stage.  Travis triggers the code involved in major events that involve integration across stages of the pipeline, such as taking the output from the forecasting stage (purple box) to create an updated presentation (rose box). Travis also runs tasks within a stage, such as conducting tests to make sure code changes have not introduced errors (icons on arrows originating and ending on the same box). ](figs/Figure1.png)

@@

A core aspect of iterative near-term forecasting is the regular rerunning of the forecasting pipeline. We chose to have the computer run our pipeline and employed "continuous analysis" [*sensu* @beaulieu2017] to drive the automation of both the full pipeline and a number of its individual components. Continuous analysis uses a set of tools originally designed for software development called "continuous integration" (CI). CI combines computing environments for running code with monitoring systems to identify changes in data or code. Essentially, CI is a computer helper whose job is to watch the pipeline and, when it sees a change in the code or data, it runs all the computer scripts needed to ensure that the forecasting pipeline runs from beginning to end. This is useful for iterative near-term forecasting because it does not rely on humans to remember to create forecasts when new models or data are added. These tools are common in the area of software development where they are used to automate software testing and integrate work by multiple developers working on the same code base. However, these tools can be used for any computational task that needs to be regularly repeated or run after changes to code or data [@beaulieu2017]. Our forecasting pipeline currently runs on a publicly available continuous integration service (Travis CI; https://travis-ci.org/) that is free for open source projects (up to a limited amount of computing time). Because of the widespread use of CI in software development, alternative services that can run code on local or cloud-based computational infrastructure also exist (e.g., Drone (http://try.drone.io/)) [@beaulieu2017]. As detailed below, we use CI to quality check data, test code using "unit tests" [@wilson2014], build models, make forecasts, and publicly present and archive the results (Figure 1b).

@@

Iterative forecasting benefits from frequently updated data so that state changes can be quickly incorporated into new forecasts [@dietze2018]. Frequent data collection and rapid processing are both important for providing timely forecasts. Since our data is collected monthly, ensuring that the models have access to the newest data requires a data latency period of less than 1 month from collection to availability for modeling. To accomplish this, we automated components of the data processing and quality assurance/quality control (QA/QC) process to reduce the time needed to add new data to the database (Figure 1).

@@

Once data is available, it needs to be processed into a form appropriate for modeling (Figure 1). For many ecological datasets, this requires not only simple data manipulation but also a good understanding of the data to facilitate appropriate aggregation. Data manipulation steps are often conducted using custom one-off code to convert the raw data into the desired form [@morris2013], but this approach has several limitations. First, each researcher must develop and maintain their own data manipulation code, which is inefficient and can result in different researchers producing different versions of the data for the same task. Subtle differences in data processing decisions have led to confusion when reproducing results for the Portal data in the past. Second, this kind of code is rarely robust to changes in data structure and location. Based on our experience developing and maintaining the Data Retriever [@morris2013; @senyondo2017], these kinds of changes are common. Finally, this kind of code is generally poorly tested, which can lead to errors based on mistakes in data manipulation. To avoid these issues for the Portal Project data, the Portal team has been developing an R package (portalR; http://github.com/weecology/portalr) for acquiring the data and handling common data cleaning and aggregation tasks. As a result, our modeling and forecasting code only needs to install this package and run the data manipulation and summary functions to get the appropriate data (Figure 1b). The package undergoes thorough automated unit tests to ensure that data manipulations are achieving the desired results. Having data manipulation code maintained in a separate package that focuses on consistently providing properly summarized forms of the most recent data has made maintaining the forecasting code itself much more straightforward.

@@

Iterative near-term forecasting involves regularly refitting a variety of different models (Figure 1). Ideally, new models should be easy to incorporate to allow for iterative improvements to the general modeling structure and approach. We use CI to refit the models and make new forecasts each time the modeling code changes and when new data becomes available (Figure 1b). We use a plugin infrastructure to allow new models to be easily added to the system. This approach treats each model as an interchangable black boxes - all models have access to the same input data and generate the same structure for model outputs (Figure 2). Each iteration of the forecasting cycle, all models that are present are run and the standardized outputs are combined into a single file to store the results of the different models' forecasts. A weighted ensemble model is then added with weights based on how well individual models fit the training data. This plugin infrastructure makes it easy to add and compare very different types of models, from the basic time-series approaches currently implemented to complex state-space and machine learning models. As long as a model script can load the provided data and produce the appropriate output it will be run and its results incorporated into the rest of the forecasting system.

![Figure 2. Demonstration of plugin infrastructure. All model scripts (represented here by the example AR1.R) are housed in a single folder. Each model script uses data provided by the core forecasting code (represented here by rodent.csv) and returns its forecast outputs in a predefined structure that is consistent across models (represented here by the example 2017_12_08forecasts.csv). Outputs from all models run on a particular date are combined into the same file (i.e. 2017_12_08forecasts.csv) to allow cross-model evaluations. Model outpus files are housed in a folder containing all forecast outputs from all previous dates to facilitate our arhciving process and and forecast assessment activities ](figs/plugin_fig.gif)

In addition to flexibility in model internals, we also wanted to support flexibility in what the models predict. Allowing models to make forecasts for system properties ranging from individual species' population abundances to total community biomass facilitates exploration of differences in forecastability across different aspects of ecological systems. We designed a forecast output format to support this. Each forecast output file contains the date being forecast, the collection date of the data used for fitting the models, the date the forecast was made, the state variable being forecast (e.g., rodent biomass, the abundance of a species), and the forecast value and associated uncertainty of that forecast (Figure 2). This allows us to store a variety of different forecasts in a common format and may serve as a useful starting point for developing a standard for storing ecological forecasts more generally.

@@

In addition to archiving the results of each forecast, we present them on a website that displays monthly rodent forecasts, model evaluation metrics, monthly reports, and information about the study site (Figure 3; [http://portal.naturecast.org](http://portal.naturecast.org/)). The website includes a graphical presentation of the most recent month's forecasts (including uncertainty) and compares the latest data to the previous forecasts. Information on the the species and the field site targeted to a general audience are also included. The site is built using Rmarkdown [@rmarkdown], which naturally integrates into the pipeline, and is automatically updated after each forecast. The `knitr` R package [@xie2015] compiles the code into HTML, which is then published using Github Pages (https://pages.github.com/). The files for the website are stored in a subdirectory of the forecasting repository. As a result, the website is also archived automatically as part of the forecast archiving.

![Figure 3. Screen capture of the homepage of the Portal Forecasting website; http://portal.naturecast.org. This site contains information on the most current forecasts, evaluation of forecast performance, and general information about the species being forecast.](figs/website_fig.png)

@@

Following the recommendations of Dietze et al [-dietze2018], we developed an automated iterative forecasting system (Figure 1) to support repeated forecasting of an ecological system. Our forecasting system automatically acquires and processes the newest data, refits the models, makes new forecasts, publicly archives those forecasts, and presents both the current forecast and information on how previous forecasts performed. Every week our forecasting system generates a new set of forecasts with no human intervention, except for the entry of new field data. This ensures that forecasts based on the most recent data are always available and allows us to rapidly assess the performance of multiple forecasting models for a number of different states of the system, including the abundances of individual species and community-level variables such as total abundance. To create this iterative near-term forecasting system, we used R to process data and conduct analyses, and we leveraged already existing services (i.e. GitHub, Travis, Docker) for more complicated cyberinfrastructure tasks. Thus, our approach to developing iterative near-term forecasting infrastructure provides an example for how short-term ecological forecasting systems can be initially developed.

We designed this forecasting system with the goal of making it relatively easy to build, maintain, and extend. We used existing technology for both running the pipeline and building individual components, which allowed us to build the system relatively cheaply in terms of both time and money. This included the use of tools like Docker for reproducibility, the Travis CI continuous integration system for automatically running the pipeline, Rmarkdown and `knitr` for generating the website, and the already existing integration between Github and Zenodo  to archive the forecasts. By using this "continuous analysis" approach [@beaulieu2017], where analyses are automatically rerun when changes are made to data, models, or associated code, we have reduced the time required by scientists to run and maintain the forecasting pipeline. To make the system extensible so that new models could be easily incorporated, we used a plugin-based infrastructure so that adding a new model to the system is as easy as adding a single file to the 'models' folder in our repository (Figure 2). This should substantially lower the barriers to other scientists contributing models to this forecasting effort. We also automatically archive the resulting forecasts publicly so that the performance of these forecasts can be assessed by both us and other researchers as new data is collected. This serves as a form of pre-registration by providing a quantitative record of the forecast before the data being predicted were collected.
